|

### 2021-12-02 06-Chp-IPD-Analysis.Rmd
End of day:

---
bibliography: bibliography/references.bib
csl: bibliography/nature.csl
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
  bookdown::word_document2: 
      toc: false
      toc_depth: 3
      reference_docx: templates/word-styles-reference-01.docx
      number_sections: false 
  bookdown::html_document2: default
documentclass: book
---

```{block type='savequote', include=knitr::is_latex_output(), quote_author='(ref:intro-quote)', echo = FALSE}
Hold
```
(ref:intro-quote) --- Hold

@@

@@

TBC

<!-- As part of a broader investigation into the relationship between lipid levels and dementia risk, I applied for access to `r n_applied$total` unique data sources containing information on these variables. However, only a small proportion of these data sources (`r fmt_applied(n_accessed$total)`) provided me with access.  -->

<!-- The resulting analysis did not suggest an effect of any lipid fraction (total cholesterol, LDL-c, HDL-c, or triglyceride) on any dementia outcome. In addition, patients characteristics such as age and sex did not have any impact on the relationship of lipid fractions and dementia outcomes. -->

<!-- Finally, the reasons for the low response rate are explored and future studies to investigate how data access rates could be improved are suggested.  -->

@@

Individual patient data (IPD) meta-analyses are considered to be the gold standard form of evidence synthesis, allowing for the application of a common selection model and analytical approach across all identified cohorts.[@riley2010] They are particularly useful for investigating the impact of patient-level characteristics, something that is not possible with aggregate data unless the results are stratified by the characteristic of interest.[@riley2010;@thompson2005] Knowledge of which groups a treatment will benefit most (or harm least) is a core aim of the move towards personalised medicine.[@riley2020;@hingorani2013]
IPD analysis also offer a mechanism by which previously unanalysed datasets can be incorporated into an analysis, thus expanding the evidence base for a particular research question.

Previous work has suggested a difference in the effect of lipids on dementia risk based on patient age and sex.[@ancelin2013; @mielke2010] While the systematic review presented in Chapter \@ref(sys-rev-heading) did not identify strong evidence for an effect of sex or age at lipid measurement on the relationship between lipids and dementia. However, the number of included studies in the lipid fraction meta-analyses was small, due to both the relatively small number of studies reporting on the lipids exposure and the poor reporting of summary statistics of patient characteristics in those that did. Additionally, best practice guidance recommends against basing the decision to perform an IPD meta-analysis on the between-study heterogeneity in a summary data meta-analysis, as similar distributions of patient covariates across studies may mask a true effect.[@riley2020]

As such, the aims of this analysis are two fold: firstly, to perform a individual patient data meta-analysis across identified cohorts to examine the impact of patient age-at-measurement and sex on the effect of lipids on dementia risk; and secondly, to expand the evidence base for the effect of lipids on dementia outcomes by obtaining estimates from previously unanalysed cohorts available via the Dementia Platform UK, a large consortia of dementia cohorts.

<!-- TODO Read this[@riley2020] -->
<!----------------------------------------------------------------------------->
&nbsp;

@@

Data sources which were cross-sectional, either by design or due to the available data (e.g. a study conducted across multiple waves, but only data from a single wave could be accessed) were excluded. Similarly, due to the limited time afforded by this thesis, studies making use of population-level electronic health records, which often require an extensive project proposals in order to gain access to the data, were ineligible due to the time and cost involved in applying. No restrictions were put on the number of participants or the length of follow-up, though participants must be free of dementia (or assumed to be free, based on age at entry)  at baseline.

@@

Where data on one of these exposures (TC, LDL-c, HDL-c, TG) was missing, it was inferred from the other three fractions (where available) using the Friedwald formula (Equation \@ref(eq:total-cholesterol-formula)). Lipids levels were retained as continuous variables rather than dichotomising into a binary hypercholesterolemia exposure, given the additional power this adds to the meta-analysis.[@riley2020;@ensor2018] Where lipid measurements were reported in _mg/dL_, these were converted to _mmol/L_. To aid interpretability of the outcome, all analyses were standardised to refer to a 1-standard deviation increase in the lipid fraction under investigation.

Across all cohorts, data cleaning was performed in a similar manner, standardising to commonly named variables so that a single model could be applied using functional programming.[@wickham2016func] The advantage of this approach is that it reduces the likelihood of errors in model mis-specification if needing to change variables names from cohort to cohort. Following data cleaning, summary statistics for each data source were calculated and compared to publicly available statistics to ensure no errors were introduced in the data cleaning process, in line with best practice.[@levis2021]

@@

With an awareness that discrepancies in available covariates is a common issue in IPD studies, I defined an idealised set of covariate domains to be age, sex, education, BMI, ApoE4 status, smoking/alcohol status and prevalent cardiovascular disease. This set covers key risk factors for dementia/Alzheimer's disease in addition to general cardiovascular risk factors. Ethnicity was considered as a potential addition covariate of interest, but several of the DPUK cohorts (whose data dictionaries could be inspected in advance) made it clear that ethnicity data was either not available or would not be shared. Details on how these variables were coded, given the available data, are presented in Section \@ref(ipd-covar-definition).

@@

Missing data in this analysis was classified as either relating to missing values (a cohort collected data on a variable of interest, but some values are missing) or missing variables (a cohort did not collect data on a variable of interest).

Variables with missing values were identified, and 20 imputed datasets were created.[@sterne2009] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) using the `mice` package in R.

Systematically missing variables were originally intended to be addressed using a previously described method.[@fibrinogenstudiescollaboration2009] Here, the correlation between the fully-adjusted and partially-adjusted estimate in cohorts with the full set of covariates is used to estimate the fully-adjusted effect in those cohorts missing covariates. However, this method requires several large cohorts to contain the full set of covariates, a condition this analysis failed to meet given the very low response rate. As such, two other common approaches were employed.[@fibrinogenstudiescollaboration2009] In the primary analysis, all cohorts were analysed adjusted for the set of common covariates across cohorts (Model 1: age, sex, smoking, alcohol, presence of morbidities). As a sensitivity analysis, cohorts with the full complement of covariates (Whitehall II and EPIC) were analysed using a maximally-adjusted model (Model 2: Model 1, further adjusted for education and BMI). Results between the common-set-adjusted (Model 1) and maximally-adjusted (Model 2) models were then compared.

@@

#### Individual patient data analysis

In this analysis, an two-stage individual patient data analysis was planned. Under a two-stage approach, estimates for the effect of each lipid fraction on available dementia outcomes were first calculated for each data source. More specifically, odds ratios were used to quantify the effect of a 1-SD lower exposure to a lipid fraction on dementia outcomes. Examination of the data available via the DPUK indicated that the common time-to-event approach used in studies of dementia outcomes would be precluded by the absence of detailed time-to-event data. An overall effect estimate was then produced by combining data-source-specific estimates in a random-effects meta-analysis (see Section \@ref(hold) in Chapter \@ref(hold) for a broader discussion of meta-analysis methods).

@@

#### Comparison with published findings

Following estimation of the main effect, where an included data source had previously been analysed, the results of this analysis were compared against those published previously. Where discrepancies were identified, these were investigated.

#### Investigating the effect of patient-level covariates

In order to investigate the interaction of patient-level characteristics (age and sex) with lipid levels, interaction terms for lipid-covariate terms were extracted and synthesised using a random effects meta-analysis.[@fisher2017] To avoid ecological bias, where the between-study association does not reflect the within-study associations, cohorts where there was no within-study variation in the covariate of interest were excluded from the interaction analysis for that covariate.[@burke2017] As an example, it is impossible to estimate the impact of sex on the lipid~dementia relationship in a study which contains female participants. All interaction analysis were performed using the common-set-adjusted model (Model 1).

<!-- QUESTION Why was Ancelin, which provides summary level data stratified by sex, not included in the analysis? Several reasons: i) cause it was not adjusted for the same covariates; ii) because the -->

@@

In summary, the requests for data from cohorts identified by the systematic review were characterised by a very low response rate. For the minority who did respond, common reasons given by authors for not sharing the data included that they: no longer worked with the same group and did not know if the data were available or how to obtain it; no longer had access to the data; or were currently performing, or intended to perform, a similar analysis as the one proposed.

<!-- TODO Need numbers in the above section. -->

With respect to the application to DPUK cohorts, where a dedicated project manager liaises with data owners on the applicants' behalf, the overall response rate was higher. However, even using this streamlined approach, a response was obtained for approximately half (N=9, 53%) of the approached cohorts, one of which (Whitehall II) was also identified by the systematic review.

<!-- TODO Fix width of this image, and one box is not rendering right, and update numbers! -->

<!----------------------------------------------------------------------------->
```{r cohortFlowchartSetup, include = FALSE}
```

@@

```{r cohortFlowchart, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:cohortFlowchart-cap)', out.width='100%', fig.scap='(ref:cohortFlowchart-scap)'}
knitr::include_graphics(file.path("figures/ipd/cohortFlowchart.png"))
```
<!----------------------------------------------------------------------------->

As highlighted in Figure \@ref(fig:cohortFlowchart), there was little overlap between cohorts identified by the systematic review and those contained in the DPUK (`r n_applied$DPUK_sysrev`), indicating that the DPUK is a useful source of unanalysed data with respect to this question.

As highlighted in Figure \@ref(fig:cohortFlowchart), nice cohorts (8 DPUK only, 1 Systematic review + DPUK) responding positively to requests for data access. However, on inspection of data provided, six of these cohorts were excluded for a range of reasons. The exclusion reason for each cohort for which data were provided is illustrated in Table \@ref(tab:dataExcluded-table). In summary, two cohorts contained were memory-complaint cohorts, containing patients unlikely to be free of dementia at baseline (BRACE, MEMENTO). Two cohorts, despite being designed as multi-wave studies, only shared data related to a single wave and so represented cross-sectional data (Generation Scotland, NICOLA). Finally TRACK HD is a genetic cohort where cohort owners advised me that dementia will inevitably develop in mid-life driven by the Huntington's (HTT gene) mutation and that this effect would likely far outweigh the effect of lipids. Finally, the ELSA study was excluded based on the dichotomous definition of the exposure variable <!-- TODO CITATION NEEDED for each study here! -->

@@

The Caerphilly Prospective Study (CaPS) is a longitudinal study of men in South Wales, UK.[@zotero-15398;@elwood2013] Blood lipids (TC, LDL-c, HDL-c and TG) were measured at baseline in 1979-1983, and from Phase III (1989-1993) onwards, a battery of cognitive tests were introduced. The cohort contains data on dementia outcomes sub-classified as vascular and non-vascular dementia. Data were available for all covariates of interest except for education, BMI and ApoE4 status.

@@

The European Prospective Investigation of Cancer (EPIC) - Norfolk is a population-based cohort, containing men and women recruited from 35 general practices in Norfolk between 1993 and 1998.[@riboli1997; @riboli2002] The added evidential value of the EPIC cohort is small, given the the fact that the data obtained contains only 8 dementia events. The cohort contains no information on dementia subtype, while all covariates of interest bar ApoE4 were available.

@@

The Whitehall II study is a prospective cohort study of men and women recruited between 1985 and 1989 from the civil service in London.[@marmot2005] The cohort is linked with the Hospital Episode Statistics (HES) database, a database containing details of patient events at NHS hospitals in England, which was used to define covariates at baseline and dementia outcomes.[@zotero-15403] Information was available on dementia type, which was classified as all-cause dementia, Alzheimer's disease or vascular dementia. This data source contained details on all covariates of interest except for ApoE4 status.

@@

Missing values in collected variables was common across the cohorts. A matrix of covariates, and whether they were not available or contained missing values for each included cohort, can be seen Appendix \@ref(hold). The Whitehall II and EPIC cohorts containted data on all but one covariates of interest, while two systematically missing covariates were identified in the CaPS cohort, namely education and BMI. No included cohort provided information on the ApoE4 status of participants, and so this variable could not be adjusted for in the analysis.

@@

Across the three cohorts, `r n_total_ipd` patients were included in this analysis (Whitehall II = 8208, EPIC = 1115, CaPS = 2512). All cohorts contained data on the four exposure variables of interest (or sufficient data from which to calculate them) and on all-cause dementia outcomes. The only other dementia outcome examined across cohorts was vascular dementia, which was reported in the CaPS and Whitehall II studies. The definitions of dementia outcomes used across cohorts can be seen in Appendix \@ref(hold). Cumulatively, there were `r n_ipd_dementia` cases of all-cause dementia, with `r n_ipd_vasdem` further classified as vascular dementia. Summary statistics for each cohort are provided in Table \@ref(tab:cohortSummary-table).

@@

@@

@@

<!-- Given minimal effect of further adjustment for education and BMI on the effect estimates, the results from the interaction analysis presented here were obtained using the common-set-adjusted model. Further, while the maximally-adjusted could have been employed in the analysis of the effect of sex, as the CaPS cohort contains only men and had to be excluded from this analysis, employing the same underlying model helped to ensure comparability between the age and sex interaction estimates.  -->

@@

@@

#### Comparison with previous analyses

For the single cohort where a previously published analysis was available (Whitehall II), results from the maximally-adjusted model for both all-cause dementia and Alzheimer's disease were comparable, with the exception of the trigluceride fraction (Figure \@ref(fig:whitehallComparison)). Tynkkynen _et al._[@tynkkynen2018] found a protective effect of triglycerides on all-cause dementia (`r estimate(0.69,0.56,0.85,"HR")`) while this analysis found evidence for a harmful effect (`r estimate(1.26,1.12,1.41,"OR")`).

<!-- TODO These are generated - check they are correct! -->

<!-- TODO Confirm the results of Tyn for HDL are different between the below figure and the results presented in the systematic review -->

<!----------------------------------------------------------------------------->
(ref:whitehallComparison-cap) Comparison of results across the two analysis of the Whitehall II cohort

(ref:whitehallComparison-scap) Comparison of results across the two analysis of the Whitehall II cohort

```{r whitehallComparison, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:whitehallComparison-cap)', out.width='100%', fig.scap='(ref:whitehallComparison-scap)'}
knitr::include_graphics(file.path("figures/ipd/whitehall_comparison.png"))
```
<!----------------------------------------------------------------------------->

@@

This analysis requested data from `r n_applied$total` data sources, but only obtained data from three, all of which were based in the United Kingdom. No evidence for an effect of lipids on the risk of dementia or related outcomes was identified, except for a harmful association of raised triglycerides with risk of vascular dementia  Similarly, there was very weak evidence for an interaction of the effect of lipids levels on dementia outcomes with participants age, grouped into 5 year bands, or sex.

Similar to the analysis of CPRD data presented in the previous chapter, this discussion will not provide a detailed comparison of the results of this analysis with other published literature, except where needed to illustrate a methodological point. For a detailed comparison of the findings presented above with the existing evidence base identified by the systematic review (Chapter \@ref(sys-rev-heading)), see Chapter \@ref(tri-heading).

@@

The obvious key limitation of this analysis is the very low response rate to requests for data access, which may bias the results if there are systematic differences between cohorts that share data and those that do not.[@ahmed2012] Whether or not to press ahead with an IPD analysis in the absence of all (or even most) data is a personal decision, and some previous analyses have highlighted where they decided not to pursue an IPD study.[@jaspers2014] For the purposes of this thesis, the decision was made to conduct the IPD analysis, as it provided training in pplication of the methods in addition to providing additional evidence sources that could be incorporated into the triangulation exercise detailed in Chapter \@ref(tri-heading).

A low response rate is not unexpected, given that a review of IPD studies published between 1987 and 2015 found that fewer than half managed to obtained data from greater than 80% of studies, and that in many cases, the exact percentage of studies for which data were obtained was not accurately reported.[@nevitt2017] However, it is assumed that a ~10% response rate is at the lower end of the scale.

There are likely several likely reasons for this low response rate to requests for data access. In general terms, there are a several well-documented reasons why data are not made readily available, including concerns regarding participant privacy and "scooping" or "parasitic" behaviour, and a lack of trust between primary and secondary researchers.[@vanpanhuis2014] More specific to this analysis, individual participant data meta-analysis including studies other than randomised controlled trials have less success in obtaining individual participant data from studies.[@nevitt2017] Additionally, while no evidence is available on whether the characteristics of the researcher requesting data access influences the response rate and eventual decision, there is the possibility that my position as a PhD student meant I was less likely to elicit the response rate than a well-known senior academic might. The timing of the requests for data access, coinciding with a global pandemic, may also have affected the response rate as researchers prioritised other COVID-related work.

@@

The barriers to data access described above are in theory what the DPUK was built to address. However, even with the help of the streamlined application process afforded by the DPUK, accessing sufficient data was a challenge. The response rate among DPUK cohorts a year after application was just 50%. In addition, some cohorts responded that that the proposed study question was already under investigation by another group, and that they would not share the data on this basis. In light of this, a centralised database of ongoing analysis being performed in DPUK would be of enormous help. Finally, the DPUK process would be aided by a clearer distinction between those cohorts that are "DPUK native" (i.e. where a copy of the data is already held on DPUK servers) versus externally hosted, as the response time for externally hosted cohorts is likely to be much longer.

@@

A key limitation of this analysis is the potential for uncontrolled confounding. Across all cohorts, adjustment for ApoE4 was not possible as I did not have access to genetic data on participants. As discussed above in relation to the Tynkkynen _et al._ analysis, ApoE4 is a strong risk factor for both increased LDL-c levels and Alzheimer's disease.[@bennet2007;@safieh2019] Failing to adjust for this factor means confounding in the LDL-c fraction results is likely.

@@

#### Discrepancy with previous analysis

I found a qualitative difference between the association of triglycerides and all-cause dementia identified in this analysis and that previously reported by Tynkkynen _et al._[@tynkkynen2018]
Investigation of this discrepancy indicated that it was likely to be due to the Whitehall snapshot that each analysis had access to, as comparison of summary statistics indicated that this analysis had substantially more dementia events (n=287 in this analysis vs. n=114 in Tynkkynen _et al._). The covariates adjusted for in each analysis were similar, though the previous analysis had access to genetic data, allowing it to adjust for ApoE4 status. However, given that ApoE4 is a risk factor for increased LDL-c rather than triglycerides,[@bennet2007] it seems unlikely that additional adjustment for this variable is responsible for the discrepancy in findings observed.

>>>>

---
bibliography: bibliography/references.bib
csl: bibliography/nature.csl
output:
  bookdown::word_document2: 
      toc: false
      toc_depth: 3
      reference_docx: templates/word-styles-reference-01.docx
      number_sections: false 
  bookdown::pdf_document2:
    template: templates/brief_template.tex
  bookdown::html_document2: default
documentclass: book
---

```{block type='savequote', include=knitr::is_latex_output(), quote_author='(ref:intro-quote)', echo = TRUE}
These particular systematic reviews [individual participant data reviews] remain yardsticks against which the quality of other reviews continues to be judged.
```
(ref:intro-quote) --- Ian Chalmers, 1993[@chalmers1993]

@@

::: {.laybox data-latex=""}

@@

<!-- TODO Finish lay summary -->

Individual participant data (IPD) meta-analyses are a special type of meta-analysis that contact the authors of relevant studies to request access to the raw data on each participant. This is different to the meta-analyses presented in the previous chapter, which used published summary results.

To examine the relationship between lipid levels and dementia risk, I applied for access to `r n_applied$total` unique data sources containing information on these variables. However, only a small proportion of these data sources (`r fmt_applied(n_accessed$total)`) provided me with access.

The resulting analysis of these data sources did not suggest a relationship between any blood lipid and any dementia outcome. The sole exception was an increased risk of vascular dementia in those with higher triglyceride levels. In addition, participants age and sex did not appear to influence the relationship of lipid fractions and dementia outcomes.

The reasons for the low response rate to request for data are explored. Finally, I suggest future studies to investigate how data access rates could be improved.

:::

@@

Individual participant data (IPD) meta-analyses are considered to be the gold standard form of evidence synthesis, allowing for the application of a common selection model and analytical approach across all identified cohorts.[@riley2010] They are particularly useful for investigating the impact of participant-level characteristics, something that is not possible with aggregate data unless the results are stratified by the characteristic of interest.[@riley2010;@thompson2005] Knowledge of which groups a treatment will benefit most (or harm least) is a core aim of the move towards personalised medicine.[@riley2020;@hingorani2013]
IPD analysis also offer a mechanism by which previously unanalysed datasets can be incorporated into an analysis, thus expanding the evidence base for a particular research question.

Previous work has suggested a difference in the effect of lipids on dementia risk based on participant age and sex,[@ancelin2013; @mielke2010] though the systematic review presented in Chapter \@ref(sys-rev-heading) did not identify strong evidence for an effect of sex or age at lipid measurement on the relationship between lipids and dementia. However, the number of included studies in the lipid fraction meta-analyses was small, due to both the relatively small number of studies reporting on the lipids exposure and the poor reporting of summary statistics of participant characteristics in those that did. Additionally, best practice guidance recommends against basing the decision to perform an IPD meta-analysis on the between-study heterogeneity in a summary data meta-analysis, as similar distributions of participant covariates across studies may mask a true effect.[@riley2020]

As such, the aims of this analysis are two fold: firstly, to perform an IPD meta-analysis across identified cohorts to examine the impact of participant age-at-measurement and sex on the effect of lipids on dementia risk; and secondly, to expand the evidence base for the effect of lipids on dementia outcomes by obtaining estimates from previously unanalysed cohorts available via the Dementia Platform UK, a large consortia of dementia cohorts.

<!-- TODO Read this[@riley2020] -->

@@

<!-- TODO Should start with inclusion criteria -->

Eligible data sources for this analysis were prospective cohort studies. Data sources which were cross-sectional, either by design or due to the available data (e.g. a study conducted across multiple waves, but only data from a single wave could be accessed) were excluded. Similarly, due to time restrictions within the scope of my thesis project, studies making use of population-level electronic health records, which often require an extensive project proposals in order to gain access to the data, were ineligible due to the time and cost involved in applying. No restrictions were put on the number of participants or the length of follow-up, though participants must be free of dementia (or assumed to be free, based on age at entry)  at baseline.

@@

Where data on one of these exposures (TC, LDL-c, HDL-c, TG) was missing, it was inferred from the other three fractions (where available) using the Friedwald formula (Equation \@ref(eq:total-cholesterol-formula)). Lipids levels were retained as continuous variables rather than dichotomising into a binary hypercholesterolemia exposure, given the additional power this adds to the meta-analysis.[@riley2020;@ensor2018] Where lipid measurements were reported in _mg/dL_, these were converted to _mmol/L_.

<!-- TODO Ensure that the results are actually on the mmol/L scale -->

Across all cohorts, data cleaning was performed in a similar manner, standardising to commonly named variables so that a single model could be applied using functional programming.[@wickham2016func] The advantage of this approach is that it reduces the likelihood of errors in model mis-specification if needing to change variables names from cohort to cohort. Following data cleaning, summary statistics for each data source were calculated and compared with publicly available statistics to ensure no errors were introduced in the data cleaning process, in line with best practice.[@levis2021]

@@

A range of additional variables were included in the analysis, intended to address the potential for confounding. With an awareness that discrepancies in available covariates is a common issue in IPD studies, I defined an idealised set of covariate domains to be age, sex, education, BMI, _ApoE4_ status, smoking/alcohol status, ethnicity, and prevalent cardiovascular disease. This set covers key risk factors for dementia/Alzheimer's disease in addition to general cardiovascular risk factors. Details on how these variables were coded, given the available data, are presented in Section \@ref(ipd-covar-definition).

@@

Missing data in this analysis was classified as either relating to missing values (a cohort collected data on a variable of interest, but some values are missing) or missing variables (a cohort did not collect data on a variable of interest). Variables with missing values were identified, and 20 imputed datasets were created.[@sterne2009] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) using the `mice` package in R.

Missing variables were originally intended to be addressed using a previously described method.[@fibrinogenstudiescollaboration2009] Here, the correlation between the fully-adjusted and partially-adjusted estimate in cohorts with the full set of covariates is used to estimate the fully-adjusted effect in those cohorts missing covariates. However, this method requires several large cohorts to contain the full set of covariates, a condition this analysis failed to meet given the very low response rate. As such, two other common approaches were employed.[@fibrinogenstudiescollaboration2009] In the primary analysis, all cohorts were analysed adjusted for the set of common covariates across cohorts (Model 1: age, sex, smoking, alcohol, presence of morbidities). As a sensitivity analysis, cohorts with the full complement of covariates (Whitehall II and EPIC) were analysed using a maximally-adjusted model (Model 2: Model 1, further adjusted for education and BMI). Results between the common-set-adjusted (Model 1) and maximally-adjusted (Model 2) models were then compared.

@@

#### IPD analysis

In this analysis, a two-stage IPD analysis was planned. Under a two-stage approach, estimates for the effect of each lipid fraction on incident dementia were first calculated for each data source. More specifically, a logistic regression model adjusted for relevant covariates as detailed in the above section was used to quantify the effect of a 1 _mmol/L_ increase in each lipid fraction on each dementia outcome. Results were expressed as odds ratios (OR). Examination of the data available via the DPUK indicated that the common time-to-event approach used in studies of dementia outcomes would be precluded by the absence of detailed time-to-event data. An overall effect estimate was then produced by combining data-source-specific estimates in a random-effects meta-analysis (see Section \@ref(hold) in Chapter \@ref(hold) for a broader discussion of meta-analysis methods).

<!-- REVIEW ORs - the one issue here may be the length of follow-up. Was it roughly equal across cohorts such that the ORs are comparable? -->

@@

#### Investigating the effect of participant-level covariates

In order to investigate the interaction of participant-level characteristics (age and sex) with lipid levels, interaction terms for lipid-covariate terms were extracted and synthesised using a fixed effects meta-analysis.[@fisher2017] To avoid ecological bias, where the between-study association does not reflect the within-study associations, cohorts where there was no within-study variation in the covariate of interest were excluded from the interaction analysis for that covariate.[@burke2017] As an example, it is impossible to estimate the impact of sex on the lipid~dementia relationship in a study which contains female participants. All interaction analysis were performed using the common-set-adjusted model (Model 1).

<!-- QUESTION   -->

@@

In summary, the requests for data from cohorts identified by the systematic review were characterised by a very low response rate (N = 5, 25%). For the minority who did respond, common reasons given by authors for not sharing the data included that they: no longer worked with the same group and did not know if the data was available or how to obtain it; no longer had access to the data; or were currently performing, or intended to perform, a similar analysis as the one proposed.

With respect to the application to DPUK cohorts, where a dedicated project manager liaises with data owners on the applicants' behalf, the overall response rate was higher. However, even using this streamlined approach, a positive response was obtained for approximately half (N=9, 53%) of the approached cohorts.

&nbsp;<!----------------------------------------------------------------------->  
```{r cohortFlowchartSetup, include = FALSE}
```

@@

```{r cohortFlowchart, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:cohortFlowchart-cap)', out.width='100%', fig.scap='(ref:cohortFlowchart-scap)'}
knitr::include_graphics(file.path("figures/ipd/cohortFlowchart.png"))
```
&nbsp;<!----------------------------------------------------------------------->

As highlighted in Figure \@ref(fig:cohortFlowchart), there was little overlap between cohorts identified by the systematic review and those contained in the DPUK (`r n_applied$DPUK_sysrev`), indicating that the DPUK is a useful source of unanalysed data with respect to this question. A single cohort (Whitehall II) was identified by the systematic review and was also present in the DPUK.

Nine cohorts (8 DPUK only, 1 Systematic review + DPUK) responded positively to requests for data access. However, on inspection of data provided, six of these cohorts were excluded and the reason for exlcusion in each case is summarised in Table \@ref(tab:dataExcluded-table). In summary, two cohorts contained were memory-complaint cohorts, containing participants unlikely to be free of dementia at baseline (BRACE, MEMENTO). Two cohorts, despite being designed as multi-wave studies, only shared data related to a single wave and so represented cross-sectional data (Generation Scotland, NICOLA). Finally TRACK HD is a genetic cohort where cohort owners advised me that dementia will inevitably develop in mid-life driven by the Huntington's (HTT gene) mutation and that this effect would likely far outweigh the effect of lipids. Finally, the ELSA study was excluded based on the dichotomous definition of the exposure variable. <!-- TODO CITATION NEEDED for each study here! -->

@@

The Caerphilly Prospective Study (CaPS) is a longitudinal study of men in South Wales, UK.[@zotero-15398;@elwood2013] Blood lipids (TC, LDL-c, HDL-c and TG) were measured at baseline in 1979-1983, and from Phase III (1989-1993) onwards, a battery of cognitive tests were introduced. The cohort contains data on dementia outcomes sub-classified as vascular and non-vascular dementia. Data was available for all covariates of interest except for education, BMI and _ApoE4_ status.

@@

The European Prospective Investigation of Cancer (EPIC) - Norfolk is a population-based cohort, containing men and women recruited from 35 general practices in Norfolk between 1993 and 1998.[@riboli1997; @riboli2002] The added evidential value of the EPIC cohort is small, given the the fact that the data obtained contains only 8 dementia events. The cohort contains no information on dementia subtype, while all covariates of interest bar _ApoE4_ were available.

@@

The Whitehall II study is a prospective cohort study of men and women recruited between 1985 and 1989 from the civil service in London.[@marmot2005] The cohort is linked with the Hospital Episode Statistics (HES) database, a database containing details of participant events at NHS hospitals in England, which was used to define covariates at baseline and dementia outcomes.[@zotero-15403] Information was available on dementia type, which was classified as all-cause dementia, Alzheimer's disease or vascular dementia. This data source contained details on all covariates of interest except for _ApoE4_ status.

@@

Missing values in collected variables was common across the cohorts. A matrix of covariates, and whether they were not available or contained missing values for each included cohort, can be seen Appendix \@ref(hold). The Whitehall II and EPIC cohorts contained data on all but one covariates of interest, while three missing covariates were identified in the CaPS cohort, namely education, BMI and ethnicity. No included cohort provided information on the _ApoE4_ status of participants, and so this variable could not be adjusted for in the analysis.

@@

Across the three cohorts, `r n_total_ipd` participants were included in this analysis (Whitehall II = 8208, EPIC = 1115, CaPS = 2512). All cohorts contained data on the four exposure variables of interest (or sufficient data from which to calculate them) and on all-cause dementia outcomes. The only other dementia outcome examined across cohorts was vascular dementia, which was reported in the CaPS and Whitehall II studies. The definitions of dementia outcomes used across cohorts can be seen in Appendix \@ref(hold). Cumulatively, there were `r n_ipd_dementia` cases of all-cause dementia, with `r n_ipd_vasdem` further classified as vascular dementia. Summary statistics for each cohort are provided in Table \@ref(tab:cohortSummary-table).

<!-- TODO Need average follow-up by cohort here - or rather put it in the table -->
<!-- TODO Need total number of Azd cases in whitehall in the paragraph above-->

@@

<!-- REVIEW Present main effect for WhiteHall AD analysis - or rather highlight that it is shown in Figure XXX in comparison with Tynken -->

@@

<!-- TODO Now that using a fixed effect model, evidence of TG/ACD association -->

@@

Given minimal effect of further adjustment for education and BMI on the effect estimates, the results from the interaction analysis presented here were obtained using the common-set-adjusted model. Further, while the maximally-adjusted could have been employed in the analysis of the effect of sex, as the CaPS cohort contains only men and had to be excluded from this analysis, employing the same underlying model helped to ensure comparability between the age and sex interaction estimates.

@@

<!-- TODO Using fixed effect meta-analyses, there is now an effect of male gender on LDL/all-cause dementia relationship. Need to discuss if sticking with fixed -->

@@

@@

This analysis requested data from `r n_applied$total` data sources, but only obtained data from three, all of which were based in the United Kingdom. No evidence for an effect of lipids on the risk of dementia or related outcomes was identified, except for a harmful association of raised triglycerides with risk of vascular dementia. Similarly, there was very weak evidence for an interaction of the effect of lipids levels on dementia outcomes with participants' age, grouped into 5 year bands, or sex.

Similar to the analysis of CPRD data presented in the previous chapter, this discussion will not provide a detailed comparison of the results of this analysis with other published literature, except to compare between this analysis and previously published results using the same data source. For a detailed comparison of the findings presented above with the existing evidence base identified by the systematic review (Chapter \@ref(sys-rev-heading)), see Chapter \@ref(tri-heading).

@@

The obvious key limitation of this analysis is the very low response rate to requests for data access, which may bias the results if there are systematic differences in the association of lipids with dementia outcomes between cohorts that share data and those that do not.[@ahmed2012] Whether or not to press ahead with an IPD analysis in the absence of all (or even most) data is a personal decision, and some previous analyses have highlighted where they decided not to pursue an IPD analysis.[@jaspers2014] For the purposes of this thesis, the decision was made to conduct the IPD analysis, as it provided training in application of the methods in addition to providing additional evidence sources that could be incorporated into the triangulation exercise detailed in Chapter \@ref(tri-heading).

A low response rate is not unexpected, given that a review of IPD studies published between 1987 and 2015 found that fewer than half managed to obtained data from greater than 80% of studies, and that in many cases, the exact percentage of studies for which data was obtained was not accurately reported.[@nevitt2017] However, it is assumed that a ~10% response rate is at the lower end of the scale.

There are likely several likely reasons for this low response rate to requests for data access. In general terms, there are a several well-documented barriers that prevent data from being made readily available, including concerns regarding participant privacy, fear of "scooping" or "parasitic" behaviour, and a lack of trust between primary and secondary researchers.[@vanpanhuis2014] More specific to this analysis, IPD meta-analysis including studies other than randomised controlled trials have less success in obtaining IPD from studies.[@nevitt2017] Additionally, while no evidence is available on whether the characteristics of the researcher requesting data access influences the response rate and eventual decision, there is the possibility that my position as a PhD student meant I was less likely to elicit the response rate than a well-known senior academic might. The timing of the requests for data access, coinciding with a global pandemic, may also have affected the response rate as researchers prioritised COVID-related work.

@@

The obstacles to data access described above are in theory what the DPUK was built to address. However, even with the help of the streamlined application process afforded by the DPUK, accessing sufficient data was a challenge. The response rate among DPUK cohorts a year after application was just 50%. In addition, some cohorts responded that that the proposed study question was already under investigation by another group, and that they would not share the data on this basis. In light of this, a centralised database of ongoing analysis being performed in DPUK would be of enormous help. Finally, the DPUK process would be aided by a clearer distinction between those cohorts that are "DPUK native" (i.e. where a copy of the data is already held on DPUK servers) versus externally hosted, as the response time for externally hosted cohorts is likely to be much longer.

@@

A key limitation of this analysis is the potential for uncontrolled confounding. Across all cohorts, adjustment for _ApoE4_ was not possible as I did not have access to genetic data on participants. _ApoE4_ is a strong risk factor for both increased LDL-c levels and Alzheimer's disease,[@bennet2007;@safieh2019] and failing to adjust for this factor means residual confounding in the LDL-c fraction results is likely.

@@

#### Comparison with previous analysis

For the single cohort (Whitehall II), a previously published analysis was available. Tynkkynen _et al._[@tynkkynen2018] analysed the association of blood lipids and risk of all-cause dementia and Alzheimer's disease across several cohorts, including the Whitehall II cohort.

In an attempt to validate my approach, I compared the results from the maximally-adjusted model (Model 2) in this analysis with those reported by Tynkkynen _et al._ for both all-cause dementia and Alzheimer's disease. The results were broadly comparable (Figures \@ref(fig:whitehallComparisonDementia) and \@ref(fig:whitehallComparisonAd)), with the exception of the association of triglycerides with all-cause dementia. Tynkkynen _et al._[@tynkkynen2018] found a protective effect of triglycerides on this outcome (`r estimate(0.69,0.56,0.85,"HR")`) while this analysis found evidence for a harmful effect (`r estimate(1.26,1.12,1.41,"OR")`).

<!----------------------------------------------------------------------------->
(ref:whitehallComparisonDementia-cap) Comparison of results across the two analysis of the Whitehall II cohort for the all-cause dementia outcome.

(ref:whitehallComparisonDementia-scap) Comparison of results across the two analysis of the Whitehall II cohort
